|

### 2022-04-15 submission/manuscript.rmd
update manuscript format for submission

---
title: |
 Individualized treatment effect was predicted best by modeling 
 baseline risk in interaction with treatment assignment
abstract: |
  **Objective**: To compare different risk-based methods for optimal prediction
  of treatment effects. **Study Design and Setting**: We simulated RCT data
  using diverse assumptions for the average treatment effect, a baseline
  prognostic index of risk (PI), the shape of its interaction with treatment
  (none, linear, quadratic or non-monotonic), and the magnitude of
  treatment-related harms (none or constant independent of the PI). We predicted
  absolute benefit using: models with a constant relative treatment effect;
  stratification in quarters of the PI; models including a linear interaction of
  treatment with the PI; models including an interaction of treatment with a
  restricted cubic spline (RCS) transformation of the PI; an adaptive approach
  using Akaike’s Information Criterion. We evaluated predictive performance
  using root mean squared error and measures of discrimination and calibration
  for benefit. **Results**: The linear-interaction model displayed optimal or
  close-to-optimal performance across many simulation scenarios with moderate
  sample size (N=4,250; ~800 events). The RCS-model was optimal for strong
  non-linear deviations from a constant treatment effect, particularly when
  sample size was larger (N=17,000). The adaptive approach also required larger
  sample sizes. These findings were illustrated in the GUSTO-I
  trial. **Conclusion**: An interaction between baseline risk and treatment
  assignment should be considered to improve treatment effect predictions.
author:
  - name: Alexandros Rekkas
    affiliation: 1
    correspondingauthor: true
  - name: Peter R. Rijnbeek
    affiliation: 1
  - name: David M. Kent
    affiliation: 2
  - name: Ewout W. Steyerberg
    affiliation: 3
  - name: David van Klaveren
    affiliation: 4
address:
  - code: 1
    address: Department of Medical Informatics, Erasmus Medical Center, Rotterdam, The Netherlands
  - code: 2
    address: Predictive Analytics and Comparative Effectiveness Center, Institute for Clinical Research and Health Policy Studies, Tufts Medical Center, Boston, Massachusetts, USA
  - code: 3
    address: Department of Biomedical Data Sciences, Leiden University Medical Center, Leiden, The Netherlands
  - code: 4
    address: Department of Public Health, Erasmus Medical Center, Rotterdam, The Netherlands
numbersections: true
linenumbers: false
journal: arXiv.org
keywords:
  - treatment effect heterogeneity
  - absolute benefit
  - prediction models
output: 
  rticles::elsevier_article:
    keep_tex: true
  bookdown::word_document2:
    reference_docx: reference.docx
geometry: margin=1.0in
date: false
toc: false
font-size: 11pt
header-includes:
  - \renewcommand*\familydefault{\sfdefault}
  - \usepackage{setspace}
  - \usepackage{amsmath}
  - \doublespacing
  - \usepackage{amssymb}
  - \usepackage{bm}
  - \usepackage{caption}
  - \usepackage{booktabs}
  - \date{}
  - \newcommand\given[1][]{\:#1\vert\:}
  - \newcommand{\indep}{\perp \!\!\! \perp}
editor_options: 
  chunk_output_type: console
bibliography: references.bib
csl: jce.csl
---

@@

\doublespacing 
\linenumbers

@@

Risk-modeling approaches to predictive HTE analyses provide a viable option in
the absence of well-established treatment effect modifier [@Kent2019;
@PathEnE]. In simulations, modeling treatment-covariate interactions, often led
to miscalibrated predictions of absolute benefit, contrary to risk-based
methods, despite their weaker discrimination of benefit in the presence of true
effect modifiers [@vanKlaveren2019]. Most often, risk-modeling approaches are
carried out in two steps: first a risk prediction model is developed externally
or internally on the entire RCT population, “blinded” to treatment; then the RCT
population is stratified using this prediction model to evaluate risk-based
treatment effect variation [@Kent2010]. This approach identified substantial
absolute treatment effect differences between low-risk and high-risk patients in
a re-analysis of 32 large trials [@Kent2016]. However, even though estimates at
the risk subgroup level may be accurate, these estimates may not apply to
individual patients.

@@

We observe RCT data $(Z, X, Y)$, where for each patient $Z_i= 0, 1$ is the
treatment status, $Y_i = 0, 1$ is the observed outcome and $X_i$ is a set of
measured covariates. Let $\{Y_i(z), z=0, 1\}$ denote the unobservable potential
outcomes. We observe $Y_i = Z_iY_i(1) + (1 - Z_i)Y_i(0)$. We are interested in
predicting the conditional average treatment effect (CATE), 
$$\tau(x) = E\{Y(0) - Y(1)|X=x\}$$ 
Assuming that $(Z, X, Y)$ is a random sample from the target population and that
$\big(Y(0), Y(1)\big)\indep Z|X$, as we are in the RCT setting, we can predict
CATE from
\begin{align*}
\tau(x) &= E\{Y(0)\given X=x\}-E\{Y(1)\given X=x\}\\
&=E\{Y\given X=x, Z=0\}-E\{Y\given X=x, Z=1\}
\end{align*}

@@

These settings resulted in a simulation study of 648 scenarios covering the
observed HTE in 32 large trials as well as many other potential variations of
risk-based treatment effect (Supplement, Sections 2 and 3) [@Kent2016].

@@

The constant treatment effect approach outperformed other approaches in the base
case scenario (N = `r 4250`; OR = 0.8; AUC= 0.75; no absolute treatment harm)
with a true constant treatment effect (median RMSE: constant treatment effect 
`r median(rmseBaseCase$constant_treatment_effect)`;
linear interaction `r median(rmseBaseCase$linear_predictor)`;
RCS-3 `r median(rmseBaseCase$rcs_3_knots)`). 
The linear interaction model was optimal under true linear deviations 
(median RMSE: constant treatment effect 
`r median(rmseLinearInteraction$constant_treatment_effect)`;
linear interaction `r median(rmseLinearInteraction$linear_predictor)`;
RCS-3 `r median(rmseLinearInteraction$rcs_3_knots)`; Figure \ref{fig:rmsebase} panels A-C)
and even in the presence of true quadratic deviations
(median RMSE: constant treatment effect 
`r median(rmseQuadratic$constant_treatment_effect)`;
linear interaction `r median(rmseQuadratic$linear_predictor)`;
RCS-3 `r median(rmseQuadratic$rcs_3_knots)`; Figure \ref{fig:rmsebase} panels A-C)
from a constant relative treatment effect. With non-monotonic deviations, RCS-3
slightly outperformed the linear interaction model
(Median RMSE: linear interaction `r median(rmseNonMonotonic$linear_predictor)`;
RCS-3 `r median(rmseNonMonotonic$rcs_3_knots)`; Figure \ref{fig:rmsebase} panel D).
With strong treatment-related harms the results were very similar in most
scenarios (Figure \ref{fig:rmsebase} panels A-C). Under non-monotonic deviations
the optimal performance of RCS-3 was more pronounced
(Median RMSE: linear interaction `r median(rmseNonMonotonicHarm$linear_predictor)`;
RCS-3 `r median(rmseNonMonotonicHarm$rcs_3_knots)`; Figure \ref{fig:rmsebase} panel D).
A stronger average treatment effect (OR=0.5) led to larger absolute benefit
predictions and consequently to larger RMSE for all approaches, but the relative
differences between different approaches were similar to the base case scenario
(Supplement, Figure S10).

@@

Increasing the sample size to `r 17000` favored RCS-3 the most. The difference in
performance with the linear interaction approach was more limited in settings
with a constant treatment effect 
(Median RMSE: linear interaction `r median(rmseBaseCaseSampleSize$linear_predictor)`;
RCS-3 `r median(rmseBaseCaseSampleSize$rcs_3_knots)`) and with a true linear interaction
(Median RMSE: linear interaction `r median(rmseLinearInteractionSampleSize$linear_predictor)`;
RCS-3 `r median(rmseLinearInteractionSampleSize$rcs_3_knots)`).
and more emphasized in settings with strong quadratic deviations 
(Median RMSE: linear interaction `r median(rmseQuadraticSampleSize$linear_predictor)`;
RCS-3 `r median(rmseQuadraticSampleSize$rcs_3_knots)`) and non-monotonic deviations
(Median RMSE: linear interaction `r median(rmseNonMonotonicSampleSize$linear_predictor)`;
RCS-3 `r median(rmseNonMonotonicSampleSize$rcs_3_knots)`).
Due to the large sample size, the RMSE of the adaptive approach was even more
similar to the best-performing method, and the constant relative treatment
effect model was less often wrongly selected (Supplement, Figure S4).

@@

Similarly, when we increased the AUC of the true prediction model to 0.85 (OR =
0.8 and N = 4,250), RCS-3 had the lowest RMSE in the case of strong quadratic or
non-monotonic deviations and very comparable performance to the – optimal –
linear interaction model in the case of strong linear deviations (median RMSE of
`r median(rmseLinearInteractionAuc$rcs_3_knots)` for RCS-3 compared to 
`r median(rmseLinearInteractionAuc$linear_predictor)` for the linear interaction
model). Similar to the base case scenario the adaptive approach wrongly selected
the constant treatment effect model
(`r adaptiveSelections %>% filter(scenarioId == 297) %>% pull(treatment) * 100`\%
and
`r adaptiveSelections %>% filter(scenarioId == 405) %>% pull(treatment) * 100`\%
of the replications in the
strong linear and non-monotonic deviation scenarios without treatment-related
harms, respectively), leading to increased variability of the RMSE (Supplement,
Figure S5).

@@

The results from all individual scenarios can be explored online at
[https://arekkas.shinyapps.io/simulation_viewer/](https://arekkas.shinyapps.io/simulation_viewer/). Additionally,
all the code for the simulations can be found at
[https://github.com/rekkasa/arekkas_HteSimulation_XXXX_2021](https://github.com/rekkasa/arekkas_HteSimulation_XXXX_2021)

@@

One limitation is that we assumed treatment benefit to be a function of baseline
risk in the majority of the simulation scenarios. We attempted to expand our
scenarios by considering constant moderate and strong treatment-related harms,
applied on the absolute scale, in line with previous work [@Glasziou1995]. In a
limited set of scenarios with true interactions between treatment assignment and
covariates, our conclusions remained unchanged (Supplement, Section 7). Even
though the average error rates increased for all the considered methods, due to
the miss-specification of the outcome model, the linear interaction model had
the lowest error rates. RCS-3 had very comparable performance. The constant
treatment effect model was often biased, especially with moderate or strong
treatment-related harms. Future simulation studies could explore the effect of
more extensive deviations from risk-based treatment effects.

>>>>

---
title: |
 Individualized treatment effect was predicted best by modeling 
 baseline risk in interaction with treatment assignment
abstract: |
author:
  - Alexandros Rekkas$^1$,
  - Peter R. Rijnbeek$^1$,
  - David M. Kent$^2$,
  - Ewout W. Steyerberg$^3$,
  - David van Klaveren$^4$
numbersections: true
linenumbers: false
output:
  bookdown::pdf_document2: default
  bookdown::word_document2:
    reference_docx: reference.docx
geometry: margin=1.0in
date: false
toc: false
font-size: 11pt
header-includes:
  - \renewcommand*\familydefault{\sfdefault}
  - \usepackage{setspace}
  - \usepackage[left, pagewise]{lineno}
  - \usepackage{caption}
  - \usepackage{amsmath}
  - \doublespacing
  - \usepackage{amssymb}
  - \usepackage{bm}
  - \usepackage{booktabs}
  - \date{}
  - \newcommand\given[1][]{\:#1\vert\:}
  - \newcommand{\indep}{\perp \!\!\! \perp}
editor_options: 
  chunk_output_type: console
bibliography: references.bib
csl: jce.csl
---

@@

\thispagestyle{empty}
\vspace{8mm}

$^1$ Department of Medical Informatics, Erasmus Medical Center, Rotterdam, The Netherlands

$^2$ Predictive Analytics and Comparative Effectiveness Center, Institute for Clinical Research and Health Policy Studies, Tufts Medical Center, Boston, Massachusetts, USA

$^3$ Department of Biomedical Data Sciences, Leiden University Medical Center, Leiden, The Netherlands

$^4$ Department of Public Health, Erasmus Medical Center, Rotterdam, The Netherlands

\vspace{10mm}
**Corresponding author**
\singlespacing 
Alexandros Rekkas, MSc

Department of Medical Informatics

Erasmus University Medical Center

3000 CA Rotterdam, P.O. Box 2040

Email: a.rekkas@erasmusmc.nl
\onehalfspacing

\vspace{10mm}
**Funding**

This work has been performed in the European Health Data and
Evidence Network (EHDEN) project. This project has received funding from the
Innovative Medicines Initiative 2 Joint Undertaking (JU) under grant agreement
No 806968. The JU receives support from the European Union’s Horizon 2020
research and innovation programme and EFPIA.

\newpage
\newpage

# Abstract {-}
**Objective**: To compare different risk-based methods for optimal prediction
of treatment effects. **Study Design and Setting**: We simulated RCT data
using diverse assumptions for the average treatment effect, a baseline
prognostic index of risk (PI), the shape of its interaction with treatment
(none, linear, quadratic or non-monotonic), and the magnitude of
treatment-related harms (none or constant independent of the PI). We predicted
absolute benefit using: models with a constant relative treatment effect;
stratification in quarters of the PI; models including a linear interaction of
treatment with the PI; models including an interaction of treatment with a
restricted cubic spline (RCS) transformation of the PI; an adaptive approach
using Akaike’s Information Criterion. We evaluated predictive performance
using root mean squared error and measures of discrimination and calibration
for benefit. **Results**: The linear-interaction model displayed optimal or
close-to-optimal performance across many simulation scenarios with moderate
sample size (N=4,250; ~785 events). The RCS-model was optimal for strong
non-linear deviations from a constant treatment effect, particularly when
sample size was larger (N=17,000). The adaptive approach also required larger
sample sizes. These findings were illustrated in the GUSTO-I
trial. **Conclusion**: An interaction between baseline risk and treatment
assignment should be considered to improve treatment effect predictions.

\vspace{10mm}

**Keywords**: treatment effect heterogeneity, absolute benefit, prediction models

\newpage

\doublespacing

@@

Risk-modeling approaches to predictive HTE analyses provide a viable option in
the absence of well-established treatment effect modifiers [@Kent2019;
@PathEnE]. In simulations, modeling treatment-covariate interactions, often led
to miscalibrated predictions of absolute benefit, contrary to risk-based
methods, despite their weaker discrimination of benefit in the presence of true
effect modifiers [@vanKlaveren2019]. Most often, risk-modeling approaches are
carried out in two steps: first a risk prediction model is developed externally
or internally on the entire RCT population, “blinded” to treatment; then the RCT
population is stratified using this prediction model to evaluate risk-based
treatment effect variation [@Kent2010]. This approach identified substantial
absolute treatment effect differences between low-risk and high-risk patients in
a re-analysis of 32 large trials [@Kent2016]. However, even though estimates at
the risk subgroup level may be accurate, these estimates may not apply to
individual patients.

@@

We observe RCT data $(Z, X, Y)$, where for each patient $Z_i= 0, 1$ is the
treatment status, $Y_i = 0, 1$ is the observed outcome and $X_i$ is a set of
measured covariates. Let $\{Y_i(z), z=0, 1\}$ denote the unobservable potential
outcomes. We observe $Y_i = Z_iY_i(1) + (1 - Z_i)Y_i(0)$. We are interested in
predicting the conditional average treatment effect (CATE), 
$$\tau(x) = E\{Y(0) - Y(1)|X=x\}$$ 
Assuming that $\big(Y(0), Y(1)\big)\indep Z|X$, as we are in the RCT setting, we
can predict CATE from
\begin{align*}
\tau(x) &= E\{Y(0)\given X=x\}-E\{Y(1)\given X=x\}\\
&=E\{Y\given X=x, Z=0\}-E\{Y\given X=x, Z=1\}
\end{align*}

@@

These settings resulted in a simulation study of 648 scenarios covering the
HTE observed in 32 large trials as well as many other potential variations of
risk-based treatment effect (Supplement, Sections 2 and 3) [@Kent2016].

@@

The constant treatment effect approach outperformed other approaches in the base
case scenario (N = `r 4250`; OR = 0.8; AUC= 0.75; no absolute treatment harm)
with a true constant treatment effect (median RMSE: constant treatment effect 
`r median(rmseBaseCase$constant_treatment_effect)`;
linear interaction `r median(rmseBaseCase$linear_predictor)`;
RCS-3 `r median(rmseBaseCase$rcs_3_knots)`). 
The linear interaction model was optimal under true linear deviations 
(median RMSE: constant treatment effect 
`r median(rmseLinearInteraction$constant_treatment_effect)`;
linear interaction `r median(rmseLinearInteraction$linear_predictor)`;
RCS-3 `r median(rmseLinearInteraction$rcs_3_knots)`; Figure \ref{fig:rmsebase} panels A-C)
and even in the presence of true quadratic deviations
(median RMSE: constant treatment effect 
`r median(rmseQuadratic$constant_treatment_effect)`;
linear interaction `r median(rmseQuadratic$linear_predictor)`;
RCS-3 `r median(rmseQuadratic$rcs_3_knots)`; Figure \ref{fig:rmsebase} panels A-C)
from a constant relative treatment effect. With non-monotonic deviations, RCS-3
slightly outperformed the linear interaction model
(median RMSE: linear interaction `r median(rmseNonMonotonic$linear_predictor)`;
RCS-3 `r median(rmseNonMonotonic$rcs_3_knots)`; Figure \ref{fig:rmsebase} panel D).
With strong treatment-related harms the results were very similar in most
scenarios (Figure \ref{fig:rmsebase} panels A-C). Under non-monotonic deviations
the optimal performance of RCS-3 was more pronounced
(median RMSE: linear interaction `r median(rmseNonMonotonicHarm$linear_predictor)`;
RCS-3 `r median(rmseNonMonotonicHarm$rcs_3_knots)`; Figure \ref{fig:rmsebase} panel D).
A stronger average treatment effect (OR=0.5) led to larger absolute benefit
predictions and consequently to larger RMSE for all approaches, but the relative
differences between different approaches were similar to the base case scenario
(Supplement, Figure S10).

@@

Increasing the sample size to `r 17000` favored RCS-3 the most (Figure
\ref{fig:rmsesamplesize}). The difference in performance with the linear
interaction approach was more limited in settings with a constant treatment
effect
(median RMSE: linear interaction `r median(rmseBaseCaseSampleSize$linear_predictor)`;
RCS-3 `r median(rmseBaseCaseSampleSize$rcs_3_knots)`) and with a true linear interaction
(median RMSE: linear interaction `r median(rmseLinearInteractionSampleSize$linear_predictor)`;
RCS-3 `r median(rmseLinearInteractionSampleSize$rcs_3_knots)`)
and more emphasized in settings with strong quadratic deviations 
(median RMSE: linear interaction `r median(rmseQuadraticSampleSize$linear_predictor)`;
RCS-3 `r median(rmseQuadraticSampleSize$rcs_3_knots)`) and non-monotonic deviations
(median RMSE: linear interaction `r median(rmseNonMonotonicSampleSize$linear_predictor)`;
RCS-3 `r median(rmseNonMonotonicSampleSize$rcs_3_knots)`).
Due to the large sample size, the RMSE of the adaptive approach was even more
similar to the best-performing method, and the constant relative treatment
effect model was less often wrongly selected (Supplement, Figure S4).

@@

Similarly, when we increased the AUC of the true prediction model to 0.85 (OR =
0.8 and N = 4,250), RCS-3 had the lowest RMSE in the case of strong quadratic or
non-monotonic deviations and very comparable performance to the – optimal –
linear interaction model in the case of strong linear deviations (median RMSE of
`r median(rmseLinearInteractionAuc$rcs_3_knots)` for RCS-3 compared to 
`r median(rmseLinearInteractionAuc$linear_predictor)` for the linear interaction
model; Figure \ref{fig:rmseauc}). Similar to the base case scenario the adaptive
approach wrongly selected the constant treatment effect model
(`r adaptiveSelections %>% filter(scenarioId == 297) %>% pull(treatment) * 100`\%
and
`r adaptiveSelections %>% filter(scenarioId == 405) %>% pull(treatment) * 100`\%
of the replications in the
strong linear and non-monotonic deviation scenarios without treatment-related
harms, respectively), leading to increased variability of the RMSE (Supplement,
Figure S5).

@@

The results from all individual scenarios can be explored online at
[https://mi-erasmusmc.shinyapps.io/HteSimulationRCT/](https://mi-erasmusmc.shinyapps.io/HteSimulationRCT/). Additionally,
all the code for the simulations can be found at
[https://github.com/mi-erasmusmc/HteSimulationRCT](https://github.com/mi-erasmusmc/HteSimulationRCT)

@@

One limitation is that we assumed treatment benefit to be a function of baseline
risk in the majority of the simulation scenarios. We attempted to expand our
scenarios by considering moderate and strong constant treatment-related harms,
applied on the absolute scale, in line with previous work [@Glasziou1995]. In a
limited set of scenarios with true interactions between treatment assignment and
covariates, our conclusions remained unchanged (Supplement, Section 7). Even
though the average error rates increased for all the considered methods, due to
the miss-specification of the outcome model, the linear interaction model had
the lowest error rates. RCS-3 had very comparable performance. The constant
treatment effect model was often biased, especially with moderate or strong
treatment-related harms. Future simulation studies could explore the effect of
more extensive deviations from risk-based treatment effects.
